 

### 2023-06-13 markdown/introduction.tex
ms explanation moved from methods to intro

---
title: " "
output: pdf_document
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
---
# Introduction

## From the Central Dogma of Molecular Biology to Omics and Systems Biology
Before understanding the underlying principle biological systems undergo and the initial exploration of molecular biology, the processes of live were roughly inferred and not well understood. The central dogma of molecular biology serves as the cornerstone of biological processes, providing a framework for understanding how genetic information is converted into functional proteins. This framework consists of two major processes, which involve transcription, converting DNA into RNA, and translation, converting RNA into functional proteins. In between those processes regulation takes base on a molecular level, by altering structure with chemical adaptations. The regulation is undertaken to provide biological organisms with an highly efficient output of their  genetic products called proteins. Crick's formulation of the central dogma has guided research in molecular biology for over 60 years \citep{Cobb, 2017}. Over the past six decades, researchers have explored various aspects of this principle, collectively referred to as omics, to comprehensively analyze biomolecules in diverse contexts. Omics research has emerged as an approach for comprehensively studying biomolecules and their interactions within biological systems. Omics encompasses various fields, including genomics, transcriptomics, proteomics, metabolomics, and others, each focusing on a specific class of biomolecules. By analyzing multiple omics layers, researchers can gain a holistic view of biological processes, uncover regulatory mechanisms, and identify key players in complex biological networks with computational methods. Systems biology takes a quantitative approach to investigate the interactions, dynamics, and emergent properties of biomolecular networks. Imagining biological systems as complex, multi-layer networks, the ensemble of proteins, known as the proteome, plays a critical role in this interactive structure with many functions such as structural integrity, catalyzing chemical reaction and regulation of cellular funcitons.

## Proteomics
The central dogma of molecular biology offers insights not only at the level of individual terms but also provides a hierarchical understanding of biology. Hierarchical organization is the underlying principle in all topics of biology. Proteins exhibit hierarchical structures that contribute to their functionality. At the primary structure level, proteins are composed of smaller building blocks called amino acids. Combining them results in the protein secondary structure such as alpha helix and beta sheet. These secondary structures further assemble to create higher-order tertiary structures, which represent specific domains within the protein. Finally, the quaternary structure describes the functional state of the protein as a whole at a given time. For instance pathogens interact with cells of an organism a cascade of biochemical reactions takes place. These reactions influence the outcome of the cellular proteome in regards of localization, abundance, post-translational modifications and more \citep{Beltran2017}. 
By studying the proteome, which encompasses all the proteins present in a biological system at a specific moment, we can gain a comprehensive understanding of the hierarchical organization and functional dynamics of biological processes.

### Bulk proteomics
Higher organisms are composed of specialized cells organized into tissues, such as skin, muscle, and blood. Each tissue consists of cells with specific functions, resulting in variations in protein expression. Bulk proteomics is a technique used to analyze the protein composition of a tissue sample, which contains all types of cells present in that particular tissue. This approach finds valuable applications for instance in oncology, where it can provide insights into the protein profile of a tumor. By studying oncogenic biomarkers and expression patterns, this knowledge can contribute to the development of screening methods and the customization of treatments based on observed expression patterns. Additionally, bulk proteomics can aid in identifying the progression of tumor development \citep{Kwon2021}. Bulk analysis expression profiles provide scientists with an average protein abundance across all cells in a sample, offering a comprehensive view of protein expression. Additionally, compared to single-cell methods, bulk samples typically contain a greater number of proteins. In fact the expanded library can be used to validate the performance of single-cell type and single-cell proteomics methods and bioinformatics pipeline \citep{Schoof2021}. However, in this type of analysis, the precise properties of individual cell types can only be inferred and remain obscured.

### Single-cell type proteomics
Taking tissue samples can lead to an averaging effect across the entire cellular ensemble, making it difficult to discern specific cell types. To overcome this limitation, cell sorting techniques such as fluorescence-activated cell sorting (FACS), magnetic-activated cell sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed \citep{Liou2015}. These techniques rely on the detection of surface proteins on cells and the specific binding of antibodies to these proteins. Antibodies, typically derived from the immune system of other species, bind to antigens in a lock-and-key manner. Antibodies are widely used in molecular biology and find applications in cell isolation. Depending on the desired application, the antibodies can be labeled with fluorescent markers (FACS), magnetic beads (MACS), or biotin (BACS). These methods have varying technical requirements, with flow cytometers being the most complex and costly. In MACS, magnetic beads are used to isolate cells from a solution by utilizing a magnet. However, mechanical forces involved in MACS can potentially damage the cells. BACS, a method developed in 2015, employs microbubbles to carry the antibodies for cellular isolation. The method showed improvements compared to MACS in regards of viability after sorting. 
It is important to note that FACS is considered the gold standard for validating the effectiveness of MACS and BACS due to its high purity and yield. FACS was used in the validation of these two methods \citep{Liou2015, Sutermaster2019}. Single-cell type proteomics reflects the protein ensemble of a specific cell type at a particular time, providing a focused perspective on the studied field compared to bulk methods \citep{Maes2020}.

### Single-cell proteomics
Cell sorting techniques primarily rely on extracellular proteins to identify and separate different cell types. However, the intracellular content of cells can still exhibit variability in experiments. Single-cell proteomics (SCP) addresses this limitation by providing a snapshot of the underlying processes within an individual cell, eliminating this variability and allowing for the elucidation of the  principles of cell type behavior, such as differentiation.
The immune system, for instance, relies on specialized cells for specific functions, and the acquisition of these functions occurs through a process called differentiation. Studying differentiation in multiple cells can be challenging due to the inherent heterogeneity during the initiation of differentiation. By analyzing one cell at a time, stimulating and observing it throughout the differentiation process, researchers can capture the entire process at a single-cell level. However, sample preparation and computational analysis in SCP can be tedious, and statistical testing can be challenging due to low cell counts.
In 2021, Specht et al. developed the SCoPE2 sample preparation and analysis pipeline, focusing on macrophage-monocyte differentiation experiments \citep{Specht2021}. This pipeline, which reduced labor time, served as a reference model for subsequent projects. SCP provides insights qualitatively and quantitatively, enabling the study of drug responses, infectious diseases, and organism development. Over the past decade, single cell techniques became indispensable tools for global research communities \citep{Minakshi2019}.

## Capturing the proteome
An early approach of qualitative analysis of the cellular proteome involved labeling with fluorescent antibodies and imaging. The major disadvantage of this technique was the limitation to only stain a few proteins per cell and the requirement of a-priori knowledge of the protein to analyze. For quantification procedures such as Western blots, immunoassays or cytometry by time of flight (CyTOF) have been used. Other disadvantages are the ability to permeate cells, accessibility and binding of the epitope and the creation of specific antibodies for a given protein \citep{Budnik2018}.
Early quantitative methods in molecular biology were assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay measuring the entire protein concentration. These methods are still in use to obtain total concentration for sample preparation. 
Another quantitative technique involved RNA-sequencing. RNAseq refers to the task of examining the transcriptome of bulk tissue sample  with deep-sequencing technologies. RNA in a sample includes mRNAs, non-coding RNAs, and small RNAs. One of the main disadvantages is, that the mapping relies on the knowledge of the DNA sequence of the observed organism (reference genome) \citep{Wang2009}. Another disadvantage is, that the amount of protein translated from the determined mRNA can only be estimated \citep{Gygi1999}. However, today these predictions can be done with computational methods such as LASSO and the accuracy of the estimate highly depends on selection and validation of this method \citep{Magnusson2022}.
When observing the protein ensemble of a cellular environment with a top-down approach  like DNAseq or RNAseq, the characteristics (e.g. post-translational modifications or charge) and local concentration of the proteins remain hidden. The cause of this phenomenon is that only a fraction of the DNA is transcribed into mRNA and not all mRNA is further translated into protein. Furthermore multiple proteins can be derived from a single DNA-sequence with alternative splicing and different proteins coming from non related sequences acting like building blocks can form protein complexes and/or Protein-RNA complexes. Cells are highly efficient and do not waste energy by producing obsolete proteins, moreover the degradation of proteins is an important energy recovery mechanism. A bottom-up strategy can elucidate these properties of the protein ensemble at a given time leading to proteomics. 
When studying impact of an external signal or stressor to a cell the abundance and   properties of a protein or protein-ensemble can enlighten the underlying pathway leading to a biological response. Since some proteins do not even function before post-translational modifications (PTMs) happen, a knowledge about these can be crucial for further studies. 
Mass spectrometry (MS) fusions the advantages of  the above mentioned techniques, because it combines the specificity of qualitative assays which are even able to sense (PTMs) with the quantitative accuracy of  protein concentration assays. Furthermore MS does not require any a-priori knowledge of the protein ensemble of the specimen, making it the method of choice for untargeted approaches.

### Mass Spectrometry
Mass spectrometry enables qualitative and quantitative analysis of the entire repertoire of a biological sample. Mass spectrometers measure the mass to charge ratio (m/z) of a particle from a fragmented larger molecule. This is achieved by a physical procedure done with a device which is made of three major components: the ion source, an analyzer, and a detector. The ion sources charges molecules and accelerate them through a magnetic field. The analyzer separates particles according to their mass to charge ratio and the detector senses charged particles and amplifies their signal \citep{Parker2010}. However before acquiring any data, sample preparation and pre-processing is required.
After sample preparation the first step involves tryptic digestion. It is a widely used technique in shotgun proteomics and involves the enzymatic cleavage of proteins into smaller peptides using the protease enzyme trypsin. This process is known as proteolysis. The enyme cleaves specifically at the caroboxyl side of arginine and lysis residues. Afterwards a charge remains on the residues making them detectable for the mass spectrometer. It is proven that proteins which are digested to the limit, meaning no further proteolysis can be induced, provide the best results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.

To observe multiple samples in one run, labeling is needed to identify each sample within a batch. 
Since mass spectrometry is not a quantitative technique by itself, the peak height or area does not reflect the abundance of a peptide. Physicochemical properties of the peptides can change the ionization efficiency and detectability of the target. However, when comparing the same analyte between multiple runs of labeled peptides, differences in the reporter ion intensity reflect the abundance of those. Labels should be chosen to change solely the mass of the sample and to not affect folding or other inherent properties of the peptide. The two techniques for labeling peptides are metabolic labeling and isobaric labling. In metabolic labeling the cells were fed with aminoacids containing heavy isotopes, it is the method of choice in order to label peptides at the earliest possible level. This atoms can be heavy nitrogen in aminoacids or salts in fertilizer for plants. Mass shifts are proportional to the isotopes incorporated during biomass production and are visible after proteolytic cleavage. Stable isotope labeling in cell culture (SILAC) was presented in the early 2000s. This method used heavy aminoacid enriched media to feed cells, in order to quantitatively analyze expression profiles. The limitation of this method is, that it can only be applied to a human cell culture (in vitro) or to model organisms (in vivo).
Isobaric labeling solely applies to in vitro techniques, with TMT and iTRAQ. The data analyzed by the developed pipeline primarily utilized TMT tags. Tandem mass tag (TMT) reagents enable differentiation of multiple samples in a single mass spectrometry (MS) run. Each sample is individually labeled with TMT reagents and then pooled together, this procedure is called multiplexing. TMTs have the same charge and differ only by their isotopic masses, the peaks observed for each sample are called reporter ions (RI). Each RI and sample is interpreted as one channel in downstream analysis. This technique allows for the identification, enrichment, and quantification of low-abundance peptide ions, which is particularly valuable in single-cell techniques. With this technique it is possible to quantify proteins and differ low abundant proteins from background noise. However, a drawback of isobaric labeling is the presence of co-fragmentation signals in the spectrogram, which requires data normalization to remove unwanted contributions \citep{Marx2019, Budnik2018}. Isotopic distribution in TMTs, reflecting the natural distribution, can be corrected during data acquisition by considering a defined spread in other channels. Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem mass tag (TMT) to enhance signal intensity of a protein species. Loading of a single-cell channel (SCC) with the right amount can be balanced to the protein concentration of a single cell, leading to single-cell proteomics \citep{Ye2022}.

Before ionization peptides need to be further separated according to their chemical properties, size or species by liquid chromatography (LC), because the number of peptide processed simultaneously by the MS device is limited.  In the case of mass spectrometry, liquid chromatography especially high performance LC (HPLC) is the technique of choice and replaces the instrumentation-wise simpler technique gel electrophoresis. In HPLC the molecules eluate through a narrow column with high pressures (50-350 bar), compared to LC where the separation is gravity dependent. Similar to gel electrophoresis molecules can be separated upon different properties depending on the type of the chromatography column. In a typical proteomics experiment peptides are separated with reversed phase chromatography (RPLC). The principle behind RPLC is a column with a nonpolar stationary phase which interacts with the nonpolar residues of the peptides. Peptides were fractionated according to their hydrophobicity.

To analyze a biological sample consisting of peptides in solution the liquid needs to be vaporized into gas phase. Two techniques are capable of this procedure. Electrospray ionization (ESI) pushes the analyte through a capillary while applying an electric current to the liquid, vaporizing the sample to a  charged aerosol. Peptides are further fragmented according to their chemical properties and can be further handled in the mass spectrometer. The fragmented peptides are now in charged droplets separated based on their surface charge, splitting further into smaller droplets until they reach the gas phase as ions. Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM). 
In the ion evaporation model the droplets shrink by evaporation until ions are expelled \citep{Iribarne1976}. The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties. In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.

Subsequently after the ionization peptides are accelerated through the magnetic field generated by the mass analyzer. Six general types of mass analyzers are availaible on the market, there are Quadropole, time of flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap and ion cyclotron resonance -mass analyzer. The function of the mass analyzer is to separate ions based on their mass to charge ratio (m/z).

Before the signal of the separated ions can be obtained an optional subsequent step involves trapping the ions an electric and/or magnetic field in order to detect ions based on their (m/z) consecutively. Ion trapping enables the technique of coupled mass spectrometers, where ions were further fragmented by collision induced dissociation and analyzed subsequently. The detector transforms the impact energy of the ion into an electric signal, called spectrum which can be processed computationally.

Given the high resolution of MS data, algorithms are employed to convert the raw signal into an interpretable form. Software packages like MaxQuant are commonly used to process the data, providing it for further analysis and statistical testing. Other software solutions include Protein Discoverer by Thermo Fisher and various R packages.

The data obtained from mass spectrometry has three dimensions: m/z ratio, intensity, and retention time. To separate peaks from each other, algorithms are used to identify local minima in the data. The centroid of each peak is determined by fitting a Gaussian peak shape, which can be interpreted as locating the peaks of each m/z spectrum as a function of retention time. The centroid of a peak corresponds to an isotope.

In order to decipher the isotopic distribution of a biomolecule, MaxQuant employs a vertex-based approach. This process, referred to as de-isotoping, utilizes graph theory. It creates a vertex for each individual peak and connects them with their potential isotopic counterparts. This is achieved by determining the mass proportion of an average amino acid to its respective isotope using a concept known as averagine \citep{Senko1995}.  By applying this procedure, the number of data points is reduced by a factor of ten, and each peak represents a small biomolecule.

The subsequent step in data acquisition involves the detection of labels for quantification purposes. Isotopic pairs of the label (such as N13, N14, N15) present in the tag or aminoacid are identified by convolving the two measured isotope patterns with the theoretical isotope patterns. Through an iterative process using a least-squares method, the best fit is determined, enabling the identification of the specific channel or sample.

The intensity-weighted average of the MS peak centroids, as described as 3D peak identification, corresponds to the mass of the peptide. The accuracy of mass measurements relies on the specific analyzer used. For an Orbitrap-type analyzer (which was used during this project), MaxQuant applies a correction value of 1 ppm (parts per million). Autocorrelation between centroids is accounted for by considering only well-identified peptides. Published data indicates that the mass precision within an MS experiment typically ranges around 10^-7.

Each peptide is represented by its individual fingerprint in the MS spectrum, which is based on the sequence and modifications of amino acids. The m/z ratio of the amino acids can be calculated, and the resulting peaks are interpreted as an amino acid sequence. However, only peptide fragments are visible in the spectrum due to peptide fragmentation. To identify proteins, peptides are matched against a sequence database \citep{Cox2008}. Sequence databases are typically stored as .fasta files and can be downloaded from the UniProt website (www.uniprot.org). The availability of gene sequences in databases and the computational matching of peptides against those sequences enable the identification of alterations in a sample at the protein level. These alterations can rely on the sequence level or could be to post-translational modifications (PTMs) such as phosphorylation, methylation \citep{Aebersold2003}. The peptide identification (P-) score reflects the goodness of fit between the data and the identified sequence in the database, taking into account the length of the peptide. This score is used to calculate the posterior error probability, which is then utilized to estimate the false discovery rate (FDR). The FDR calculation provides a measure of the rate at which false identifications are expected in the dataset.

After these calculations, the identified peptides can be grouped and aggregated according to their respective proteins for quantification. Various metrics that indicate the performance of the peptide search can be utilized in downstream analysis. Within the identified peptide groups, a "razor protein" refers to the protein with the highest number of ambiguously identified peptides. Quantification is performed by considering only unique peptides for comparison. Posterior error probabilities, which represent the likelihood of the identified peptide being a random event, are multiplied together, and only distinct sequences with the highest scores are considered for further analysis.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{float}
\usepackage{sectsty}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}{%
\subsection{From the Central Dogma of Molecular Biology to Omics and
Systems
Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}

Before understanding the underlying principle biological systems undergo
and the initial exploration of molecular biology, the processes of live
were roughly inferred and not well understood. The central dogma of
molecular biology serves as the cornerstone of biological processes,
providing a framework for understanding how genetic information is
converted into functional proteins. This framework consists of two major
processes, which involve transcription, converting DNA into RNA, and
translation, converting RNA into functional proteins. In between those
processes regulation takes base on a molecular level, by altering
structure with chemical adaptations. The regulation is undertaken to
provide biological organisms with an highly efficient output of their
genetic products called proteins. Crick's formulation of the central
dogma has guided research in molecular biology for over 60 years
\citep{Cobb2017}. Over the past six decades, researchers have explored
various aspects of this principle, collectively referred to as omics, to
comprehensively analyze biomolecules in diverse contexts. Omics research
has emerged as an approach for comprehensively studying biomolecules and
their interactions within biological systems. Omics encompasses various
fields, including genomics, transcriptomics, proteomics, metabolomics,
and others, each focusing on a specific class of biomolecules. By
analyzing multiple omics layers, researchers can gain a holistic view of
biological processes, uncover regulatory mechanisms, and identify key
players in complex biological networks with computational methods.
Systems biology takes a quantitative approach to investigate the
interactions, dynamics, and emergent properties of biomolecular
networks. Imagining biological systems as complex, multi-layer networks,
the ensemble of proteins, known as the proteome, plays a critical role
in this interactive structure with many functions such as structural
integrity, catalyzing chemical reaction and regulation of cellular
funcitons.

\hypertarget{proteomics}{%
\subsection{Proteomics}\label{proteomics}}

The central dogma of molecular biology offers insights not only at the
level of individual terms but also provides a hierarchical understanding
of biology. Hierarchical organization is the underlying principle in all
topics of biology. Proteins exhibit hierarchical structures that
contribute to their functionality. At the primary structure level,
proteins are composed of smaller building blocks called amino acids.
Combining them results in the protein secondary structure such as alpha
helix and beta sheet. These secondary structures further assemble to
create higher-order tertiary structures, which represent specific
domains within the protein. Finally, the quaternary structure describes
the functional state of the protein as a whole at a given time. By
studying the proteome, which encompasses all the proteins present in a
biological system at a specific moment, we can gain a comprehensive
understanding of the hierarchical organization and functional dynamics
of biological processes.

\hypertarget{bulk-proteomics}{%
\subsubsection{Bulk proteomics}\label{bulk-proteomics}}

Higher organisms are composed of specialized cells organized into
tissues, such as skin, muscle, and blood. Each tissue consists of cells
with specific functions, resulting in variations in protein expression.
Bulk proteomics is a technique used to analyze the protein composition
of a tissue sample, which contains all types of cells present in that
particular tissue. This approach finds valuable applications for
instance in oncology, where it can provide insights into the protein
profile of a tumor. By studying oncogenic biomarkers and expression
patterns, this knowledge can contribute to the development of screening
methods and the customization of treatments based on observed expression
patterns. Additionally, bulk proteomics can aid in identifying the
progression of tumor development \citep{Kwon2021}. Bulk analysis
expression profiles provide scientists with an average protein abundance
across all cells in a sample, offering a comprehensive view of protein
expression. Additionally, compared to single-cell methods, bulk samples
typically contain a greater number of proteins. In fact the expanded
library can be used to validate the performance of single-cell type and
single-cell proteomics methods and bioinformatics pipeline
\citep{Schoof2021}. However, in this type of analysis, the precise
properties of individual cell types can only be inferred and remain
obscured.

\hypertarget{single-cell-type}{%
\subsubsection{Single-cell type}\label{single-cell-type}}

Taking tissue samples can lead to an averaging effect across the entire
cellular ensemble, making it difficult to discern specific cell types.
To overcome this limitation, cell sorting techniques such as
fluorescence-activated cell sorting (FACS), magnetic-activated cell
sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed
\citep{Liou2015}. These techniques rely on the detection of surface
proteins on cells and the specific binding of antibodies to these
proteins. Antibodies, typically derived from the immune system of other
species, bind to antigens in a lock-and-key manner. Antibodies are
widely used in molecular biology and find applications in cell
isolation. Depending on the desired application, the antibodies can be
labeled with fluorescent markers (FACS), magnetic beads (MACS), or
biotin (BACS). These methods have varying technical requirements, with
flow cytometers being the most complex and costly. In MACS, magnetic
beads are used to isolate cells from a solution by utilizing a magnet.
However, mechanical forces involved in MACS can potentially damage the
cells. BACS, a method developed in 2015, employs microbubbles to carry
the antibodies for cellular isolation. The method showed improvements
compared to MACS in regards of viability after sorting. It is important
to note that FACS is considered the gold standard for validating the
effectiveness of MACS and BACS due to its high purity and yield. FACS
was used in the validation of these two methods
\citep{Liou2015, Sutermaster2019}. Single-cell type proteomics reflects
the protein ensemble of a specific cell type at a particular time,
providing a focused perspective on the studied field compared to bulk
methods \citep{Maes2020}.

\hypertarget{single-cell-proteomics}{%
\subsubsection{Single-cell proteomics}\label{single-cell-proteomics}}

Cell sorting techniques primarily rely on extracellular proteins to
identify and separate different cell types. However, the intracellular
content of cells can still exhibit variability in experiments.
Single-cell proteomics (SCP) addresses this limitation by providing a
snapshot of the underlying processes within an individual cell,
eliminating this variability and allowing for the elucidation of the
underlying principles of cell type behavior, such as differentiation.
The immune system, for instance, relies on specialized cells for
specific functions, and the acquisition of these functions occurs
through a process called differentiation. Studying differentiation in
multiple cells can be challenging due to the inherent heterogeneity
during the initiation of differentiation. By analyzing one cell at a
time, stimulating and observing it throughout the differentiation
process, researchers can capture the entire process at a single-cell
level. However, sample preparation and computational analysis in SCP can
be tedious, and statistical testing can be challenging due to low cell
counts. In 2021, Specht et al.~developed the SCoPE2 sample preparation
and analysis pipeline, focusing on macrophage-monocyte differentiation
experiments \citep{Specht2021}. This pipeline, which reduced labor time,
served as a reference model for subsequent projects. SCP provides
insights qualitatively and quantitatively, enabling the study of drug
responses, infectious diseases, and organism development. Over the past
decade, single cell techniques became indispensable tools for global
research communities \citep{Minakshi et al., 2019}.

An early approach of qualitative analysis of the cellular proteome
involved labeling with fluorescent antibodies and imaging. The major
disadvantage of this technique was the limitation to only stain a few
proteins per cell. For quantification procedures such as single-cell
Western blots, immunoassays or CyTOF have been used. Other disadvantages
are the ability to permeate cells, accessibility and binding of the
epitope and the creation of specific antibodies for a given protein
\citep{Budnik2018}.

One of those techniques involved RNA-sequencing. Since RNA involves also
non-coding RNA, the amount of RNA is often not proportional to the
content of proteins in a cell. So the proteinaceous content of a cell
was only predicted and quantitative analysis was not possible.

\hypertarget{effectors-of-the-proteome}{%
\subsection{Effectors of the proteome}\label{effectors-of-the-proteome}}

\hypertarget{external-effectors}{%
\subsubsection{External effectors}\label{external-effectors}}

When pathogens interact with cells of an organism a cascade of
biochemical reactions takes place. These reactions influence the outcome
of the cellular proteome in regards of localization, abundance,
post-translational modifications and more (Jean Beltran et al., 2017).

\hypertarget{methods-for-quantifications}{%
\subsection{Methods for
Quantifications}\label{methods-for-quantifications}}

A holistic approach to analyze the expression can be sequencing of RNA
(RNAseq) or mass-spectrometry (MS). RNAseq refers to the task of
examining the transcriptome of bulk tissue sample with deep-sequencing
technologies. RNA in a sample includes mRNAs, non-coding RNAs, and small
RNAs. One of the main disadvantages is, that the mapping relies on the
knowledge of the DNA sequence of the observed organism (reference
genome) (Wang et al., 2009). Another disadvantage is, that the amount of
protein translated from the determined mRNA can only be estimated (Gygi
et al., 1999). However, today these predictions can be done with
computational methods such as LASSO and the accuracy of the estimate
highly depends on selection and validation of this method (Magnusson et
al., 2022).

When observing the protein ensemble of a cellular environment with a
top-down approach like DNAseq or RNAseq, the characteristics
(e.g.~post-translational modifications or charge) and local
concentration of the proteins remain hidden. The cause of this
phenomenon is that only a fraction of the DNA is transcribed into mRNA
and not all mRNA is further translated into protein. Furthermore
multiple proteins can be derived from a single DNA-sequence with
alternative splicing and different proteins coming from non related
sequences acting like building blocks can form protein complexes and/or
Protein-RNA complexes. Cells are highly efficient and do not waste
energy by producing obsolete proteins, moreover the degradation of
proteins is an important energy recovery mechanism. A bottom-up strategy
can elucidate these properties of the protein ensemble at a given time
leading to proteomics. Early methods in molecular biology were western
blots with the disadvantage of having a-priori knowledge of the analyzed
protein and quantitative protein assays like the UV-vis-, Bradford-
and/or bicinchoninic-acid-assay measuring the entire protein
concentration. When studying impact of an external signal or stressor to
a cell the abundance and properties of a protein or protein-ensemble can
enlighten the underlying pathway leading to a biological response. Since
some proteins do not even function before post-translational
modifications happen, a knowledge about these can be crucial for further
studies.

\hypertarget{mass-spectrometry}{%
\subsubsection{Mass Spectrometry}\label{mass-spectrometry}}

Mass spectrometry enables qualitative and quantitative analysis of the
entire repertoire of a biological sample. The availability of gene
sequences in databases and the ability to match proteins against those
sequences with computational methods makes it possible to identify
alterations of a sample on a protein level. These alterations can rely
on the sequence level or could be to post-translational modifications
(PTMs) such as phosphorylation, methylation or else
\citep{Aebersold2003}.

Mass spectrometers measure the mass to charge ratio (m/z) of a particle
from a fragmented larger molecule. This is achieved by a physical
procedure done with a device which is made of three major components:
the ion source, an analyzer, and a detector. The ion sources charges
molecules and accelerate them through a magnetic field. The analyzer
separates particles according to their mass to charge ratio and the
detector senses charged particles and amplifies their signal (Parker et
al., 2010). MS provides great sensitivity for traces, hence getting an
average concentration for all biomolecules in a sample. To identify
protein abundance from a single type of cell, cell sorting techniques
needed to be improved and applied to MS (Minakshi et al., 2019).
Single-cell proteomics by mass spectrometry (SCoPE-MS) uses a isobaric
labeling with a tandem mass tag (TMT) to enhance signal intensity of a
protein species. Loading of a single-cell channel (SCC) with the right
amount can be balanced to the protein concentration of a single cell,
leading to single-cell proteomics (Ye et al., 2022).

MS provides great sensitivity for traces, hence getting an average
concentration for all biomolecules in a sample. To identify protein
abundance from a single type of cell, cell sorting techniques needed to
be improved and applied to MS (Minakshi et al., 2019). Single-cell
proteomics by mass spectrometry (SCoPE-MS) uses a isobaric labeling with
a tandem mass tag (TMT) to enhance signal intensity of a protein
species. Loading of a single-cell channel (SCC) with the right amount
can be balanced to the protein concentration of a single cell, leading
to single-cell proteomics (Ye et al., 2022).

\end{document}
